---
title: "R-Studio: Simple Linear Regression"
subtitle: HW3
output: ioslides_presentation
---

<style type="text/css">
body p, div,h3, h4, h5 {
color: black;
font-family: 'Karla',sans-serif;
font-size: 18px;
}
slides > slide.title-slide hgroup h1 {
font-family:'Montserrat', sans-serif;
color: #597B8D;
}

h2 {
font-family: 'Montserrat', sans-serif;
color: #597B8D;
}

pre {
  font-size: 18px; /* Adjust the font size as needed */
}

</style> 
<!-- end of defining font in various parts of slides -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is Linear Regression?

Linear Regression is a statistical technique that aims at measuring the relationship between two variables. It helps to visually see the trend and relationships via a 'line of best fit' in graphs. As well as allowing predictions of the dependent variable for any given independent variable. <br>
The equation is derived from this formula: <br>
$$ y = mx+b $$ <br>
where: <br>
$y$ = dependent variable <br>
$x$ = independent variable <br>
$m$ = slope <br>
$b$ = y-intercept <br>


## Linear Regression Equation

$$ y = \beta_0 + \beta_1 \cdot x + \varepsilon $$
where: <br>
$y$ = dependent variable <br>
$\beta_0$ = intercept <br>
$\beta_1$ = slope <br>
$x$ = independent variable <br>
$\varepsilon$ = error term <br>

## Visualizing Linear Regression in R-Studio: Data Import

For this example we will be using the palmerpenguins data set. Collected by Dr.Kristen Gorman, this data set contains information for 344 penguins, across 3 different species and 3 islands in the Palmer Archipelago.

```{r cars, echo = TRUE, message = FALSE, warning=FALSE}

library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(plotly)
data(penguins)


```

## Visualizing Linear Regression in R-Studio: Constructing Plot

Code to create simple scatter plot:

```{r, echo = TRUE, eval = FALSE , warning = FALSE, message = FALSE}
ggplot(penguins, aes(x = flipper_length_mm,
                          y = bill_length_mm,
                          color = species))+
  geom_point() +
  labs(x = "Flipper Length (in mm)",
       y = "Bill Length (in mm)",
       color = "Species",
       title = "Comparing Penguin Bill to Flipper Length")
```

## Visualizing Linear Regression in R-Studio: Constructing Plot Cont.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
ggplot(penguins, aes(x = flipper_length_mm,
                          y = bill_length_mm,
                          color = species))+
  geom_point() +
  labs(x = "Flipper Length (in mm)",
       y = "Bill Length (in mm)",
       color = "Species",
       title = "Comparing Penguin Bill to Flipper Length")
```


## Visualizing Linear Regression in R-Studio: Adding Regression Line
`geom_smooth()` adds a 'line of best fit'. If you do not state a `method()`
then by default it will plot a smooth curve. In order to get a line you 
must declare `geom_smooth(method = "lm")`.
```{r, echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE}
#---------- Constructs basic scatter plot ------------
ggplot(penguins, aes(x = flipper_length_mm,
                     y = bill_length_mm,
                     color = species)) +
  geom_point() +
  # Adds regression line
  geom_smooth(method = "lm",
              se = FALSE) +
  # Adds labels 
  labs(x = "Flipper Length (in mm)",
       y = "Bill Length (in mm)",
       color = "Species",
       title = "Comparing Penguin Bill to Flipper Length")
```


## Visualizing Linear Regression in R-Studio: Adding Regression Line Cont.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
#---------- Constructs basic scatter plot ------------
ggplot(penguins, aes(x = flipper_length_mm,
                     y = bill_length_mm,
                     color = species)) +
  geom_point() +
  # Adds regression line
  geom_smooth(method = "lm",
              se = FALSE) +
  # Adds labels 
  labs(x = "Flipper Length (in mm)",
       y = "Bill Length (in mm)",
       color = "Species",
       title = "Comparing Penguin Bill to Flipper Length")

```


## Interactive  3D Scatter Plot with Plotly

```{r}
penguins_clean <- na.omit(penguins)

x <- penguins_clean$flipper_length_mm
y <- penguins_clean$bill_length_mm
z <- penguins_clean$bill_depth_mm
color <- penguins_clean$species

xax <- list(title = "Flipper Length (in mm)",
            titlefont = list(size = 10))
yax <- list(title = "Bill Length ( in mm",
            titlefont = list(size = 10))
zax <- list(title = "Bill Depth ( in mm",
            titlefont = list(size = 10))

# Create Plotly scatter plot
fig <- plot_ly(data = penguins_clean,
               x = x,
               y = y,
               z = z,
               type = "scatter3d",
               mode = "markers",
               color = ~color,
               marker = list(size = 4)) %>%
       layout( scene = list( 
         xaxis = xax,
         yaxis = yax,
         zaxis = zax 
         ),
         title = "Penguin Data: 3D Scatter Plot"
        )
fig

```


